DCMIP2016, Part 1: Models and Equation Sets

### 2017-04-24 DCMIP2016-Part1-Models.tex
- Merged updates from FVM

This paper is the first in a series of papers documenting the results of this workshop.  It is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling.  For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.  The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  The summary and conclusions then follow in section \ref{sec:Conclusions}.  Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.

@@

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH, D & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, {\color{red} $q$} & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & fixed height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
FVM & Monotonic limiting & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The horizontal placement of variables impacts a number of properties of the numerical method, including the ability of the model to conserve invariant such as energy and enstrophy, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.  The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.  Later other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.  New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.  Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).

@@

%----------------------------------------------------
\begin{figure}[]
\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.}
\label{fig:horizontalstaggering}
\end{figure}
%----------------------------------------------------

@@

In the floating Lagrangian formulation \citep{lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.  The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
\begin{align}
s = \int^{\infty}_{z} \rho_0 g dz,
\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
\begin{align}
\dot{s} = \frac{ds}{dt} = 0,
\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.  Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.

@@

The vast diversity of the modern dynamical core ecosystem suggests that there is no singular approach that is intrinsically superior to other methods.  Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.  The repercussions that emerge from these choices can then be explored in the context of idealized test cases, such as the ones that have been proposed as part of DCMIP.  The remaining papers in this series investigate how the models described in this paper are able to simulate three idealized test cases, which each incorporate simplified model physics:  a moist baroclinic wave, an idealized tropical cyclone, and a splitting supercell storm on a small planet.  Where appropriate, metrics have been included that may be indicative of model performance.  These tests can also be used for future dynamical core development to identify where a new dynamical core diverges from a suite of modern models.

>>>>

This paper is the first in a series of papers documenting the results of this workshop.  Its purposes is twofold:  First, to review the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling; and second, to provide a mechanism to understand the differences that arise in the test cases of later papers in this series.  For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models, along with a tabulation of relevant details about the dynamical core design.  The body of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:VariablePlacement} describes aspects of variable placement and prognosis; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  The summary and conclusions then follow in section \ref{sec:Conclusions}.  Finally, appendix \ref{sec:EquationSets} provides a comprehensive overview of the various forms the Navi\'er-Stokes equations take in dynamical cores, and has been included as a resource for dynamical core developers.

@@

\begin{table}[p]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or non-hydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)?} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\delta \Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\vb{u}_h$, $w$, $\theta'$, $q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, {\color{red}$\log \partial s / \partial \zeta$}, $\log T_v$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $\vb{u}_h$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
OLAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\theta_{il}$, {\color{red} $q$} & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Vertical staggering (detailed in section \ref{sec:VerticalStaggering}) and vertical coordinates (detailed in section \ref{sec:VerticalCoordinates}) for participating dynamical cores.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Staggering & Vertical Coordinate \\ \hline 
CSU & Lorenz & fixed height \\
DYNAMICO & Lorenz & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FV$^3$ & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
FVM & Co-located & fixed height \\
GEM & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering}) & log pressure (\S \ref{sec:GEM_zeta})  \\
HOMME & Co-located & floating mass (\S \ref{sec:FloatingLagrangian}) \\
ICON & Lorenz & fixed height \\
MPAS & Lorenz & fixed height \\
NICAM & Lorenz & fixed height \\
OLAM & Lorenz & fixed height with cut-cells (\S \ref{sec:CutCells}) \\
Tempest & Lorenz & fixed height \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[p]
\caption{Standard options for diffusion, stabilization, filters, or fixers in participating dynamical cores (detailed in section \ref{sec:DiffusionStabilization}) and temporal discretization (detailed in section \ref{sec:TemporalDiscretizations}).} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperviscosity & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, hyperviscosity & Forward-backwards \citep{LR1997QJR} / semi-implicit \\
FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & 4th-order horizontal hyperviscosity & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
OLAM & Divergence/vorticity damping & 2nd-order Adams-Bashforth, Lax-Wendroff (for tracers) \\
Tempest & 4th-order horizontal hyperviscosity & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The horizontal placement of variables impacts a number of properties of the numerical method, including how energy and enstrophy conservation is managed, any computational modes that might arise due to differencing, and the dispersion properties of the method \citep{randall1994geostrophic}, and the maximum stable timestep size \citep{ullrich2014understanding}.  The original four Arakawa grids \citep{arakawa1977computational}, denoted with letters A- through D- were initially designed for rectilinear meshes, but were later adapted for a variety of unstructured grids.  Later, other grid types were added, including the Z-grid, which used the vertical component of vorticity and the horizontal divergence in place of the velocity components \citep{randall1994geostrophic}, and the ZM-grid, which extends the B-grid to hexagons by placing the velocity at hexagonal nodes \citep{ringler2002zm}.  New staggerings are now being developed in the context of finite element methods, using mixed finite element methods \citep{cotter2012mixed}.  Among the models that participated in DCMIP, only four grids were represented:  The A-grid, which involves co-location of all velocity components; the C-grid, which places perpendicular velocity components on grid edges; the D-grid, which places parallel velocity components on grid edges; and the Z-grid (see Figure \ref{fig:horizontalstaggering}).

@@

%----------------------------------------------------
\begin{figure}[]
\includegraphics[width=5in, clip, trim=0cm 2.0cm 0cm 0cm]{HorizontalStaggering.pdf}
\caption{Horizontal staggering options represented among DCMIP models, in this case depicted on a rectilinear grid and geodesic grid.  Here $\eta$ denotes the buoyancy variable.}
\label{fig:horizontalstaggering}
\end{figure}
%----------------------------------------------------

@@

In the floating Lagrangian formulation \citep{starr1945quasi, lin2004vertically} the vertical coordinate is chosen to represent an artificial tracer with monotonically increasing or decreasing mixing ratio $s$ in the vertical.  The actual mixing ratio at initiation is arbitrary, and can be constructed to be height-like (i.e., $s = z$), or mass-like, i.e.
\begin{align}
s = \int^{\infty}_{z} \rho_0 g dz,
\end{align} in which case a 3D reference density field $\rho_0$ can be imposed.  Of primary importance is the fact that the vertical coordinate satisfy
\begin{align}
\dot{s} = \frac{ds}{dt} = 0,
\end{align} which greatly simplifies the associated prognostic velocity and continuity equations.  Floating Lagrangian coordinates are often paired with a vertical remapping operation that corrects for strong grid distortions that may occur after sufficiently long model integrations.

@@

The vast diversity of the modern dynamical core ecosystem suggests that there is no consensus on a single approach that is intrinsically superior to other options.  Choices made in the dynamical core confer advantages that include parallel scalability \citep{dennis2012cam}, conservation of invariants \citep{thuburn2008some}, or representation of the Kinetic energy spectrum \citep{skamarock2004evaluating}.  The repercussions that emerge from these choices can then be explored in the context of idealized test cases, such as the ones that have been proposed as part of DCMIP.  The remaining papers in this series investigate how the models described in this paper are able to simulate three idealized test cases, which each incorporate simplified model physics:  a moist baroclinic wave, an idealized tropical cyclone, and a splitting supercell storm on a small planet.  Where appropriate, metrics have been included that may be indicative of model performance.  These tests can also be used for future dynamical core development to identify where a new dynamical core diverges from a suite of modern models.
